\bf Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2020-09-11 MainDocument/MainDocument.Rmd
New small section on Abductive reasoning.

>>>>

# Abductive Reasoning and Causal Inference

Scientific inquiry falls under one of three modes of logical reasoning: deduction, induction, and abduction. Deductive reasoning starts with a theory and makes a prediction of what the observations should be if that theory is correct. The classic example of deductive reasoning essentially goes (a) "all men are mortal", (b) "Socrates is a man", (c) "Therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning. For example, John Snow used deductive reasoning in to stop the 1854 London cholera outbreak. He hypothesized that cholera was waterborne and deduced that a bad water pump was responsible for the outbreak.

Inductive reasoning is the opposite of deductive reasoning. Rather than starting with a theory and making a prediction, induction starts with an observation and makes a theory based on the likeliest reason. For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that hispanics in the United States had, paradoxically, better health outcomes than expected given their socioeconomic status. Based on this observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science but are not without their shortcomings. In both approaches, it is impossible, or at least imporbable, for the premise to be true and the conclusion false. Furthermore, if a hypothesis is rejected under a deductive or inductive approach, there is no alternative explanation -- there is either the presence or absence of support.

Abductive reasoning is a third, less well-known way that reasons the best explanation given the observations [@walton2014abductive]. It is most concerned with generating and evaluating explanatory hypotheses [@Crowder2017] by looking for a pattern and suggesting a hypothesis [@fann2012peirce;@pierce1878deduction]. Unlike the more rigid deductive and inductive reasoning, abductive reasoning allows the scientist to move between deduction and induction and fits well during exploratory data analysis [@yu1994abduction;@haig2015commentary;@tukey1977exploratory]. During exploratory data analysis, abductive reasoning helps find a pattern and suggest a hypothesis, deduction helps refine and build a testable hypothesis, and induction is the empirical support [@yu1994abduction].

Abductive reasoning and exploratory data analysis are best employed in a data rich environment -- the very data environment social scientists and demographers find themselves in the present moment. Additionally, in a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020]because it focuses our attention on only that which we seek^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery? They created a simple dataset that if plotted would reveal an image of a gorilla waving. Students instructed to find support for specific hypotheses were less likely to discover the gorilla compared to those who were simply asked "What do you conclude from the dataset?"]. Focusing on specific hypotheses often prevents us from finding other interesting phenomena. Afterall, it was the publication of the hispanic paradox itself that has generated more than thirty years of devoted scholarship. When paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results in the absence of hypothesis testing.

### 2020-09-11 MainDocument/MainDocument.tex
New small section on Abductive reasoning.

>>>>

\hypertarget{abductive-reasoning-and-causal-inference}{%
\section{Abductive Reasoning and Causal
Inference}\label{abductive-reasoning-and-causal-inference}}

Scientific inquiry falls under one of three modes of logical reasoning:
deduction, induction, and abduction. Deductive reasoning starts with a
theory and makes a prediction of what the observations should be if that
theory is correct. The classic example of deductive reasoning
essentially goes (a) ``all men are mortal'', (b) ``Socrates is a man'',
(c) ``Therefore, Socrates is mortal.'' Deductive reasoning has a long
history in the sciences and most hypothesis testing in the social
sciences follows this logical reasoning. For example, John Snow used
deductive reasoning in to stop the 1854 London cholera outbreak. He
hypothesized that cholera was waterborne and deduced that a bad water
pump was responsible for the outbreak.

Inductive reasoning is the opposite of deductive reasoning. Rather than
starting with a theory and making a prediction, induction starts with an
observation and makes a theory based on the likeliest reason. For
example, the Hispanic Mortality Paradox first appeared in the scientific
literature in the mid-1980s \citep{markides1986health} as an observation
that hispanics in the United States had, paradoxically, better health
outcomes than expected given their socioeconomic status. Based on this
observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science
but are not without their shortcomings. In both approaches, it is
impossible, or at least imporbable, for the premise to be true and the
conclusion false. Furthermore, if a hypothesis is rejected under a
deductive or inductive approach, there is no alternative explanation --
there is either the presence or absence of support.

Abductive reasoning is a third, less well-known way that reasons the
best explanation given the observations \citep{walton2014abductive}. It
is most concerned with generating and evaluating explanatory hypotheses
\citep{Crowder2017} by looking for a pattern and suggesting a hypothesis
\citep{fann2012peirce, pierce1878deduction}. Unlike the more rigid
deductive and inductive reasoning, abductive reasoning allows the
scientist to move between deduction and induction and fits well during
exploratory data analysis
\citep{yu1994abduction, haig2015commentary, tukey1977exploratory}.
During exploratory data analysis, abductive reasoning helps find a
pattern and suggest a hypothesis, deduction helps refine and build a
testable hypothesis, and induction is the empirical support
\citep{yu1994abduction}.

Abductive reasoning and exploratory data analysis are best employed in a
data rich environment -- the very data environment social scientists and
demographers find themselves in the present moment. Additionally, in a
data rich environment, a hypothesis is a hidden liability
\citep{yanai_hypothesis_2020}because it focuses our attention on only
that which we seek\footnote{Yanai and Lercher
  \citeyearpar{yanai_hypothesis_2020} ask if focusing on a specific
  hypothesis prevents new discovery? They created a simple dataset that
  if plotted would reveal an image of a gorilla waving. Students
  instructed to find support for specific hypotheses were less likely to
  discover the gorilla compared to those who were simply asked ``What do
  you conclude from the dataset?''}. Focusing on specific hypotheses
often prevents us from finding other interesting phenomena. Afterall, it
was the publication of the hispanic paradox itself that has generated
more than thirty years of devoted scholarship. When paired with causal
inference techniques, abductive reasoning provides an empirical platform
for the discovery of potentially interesting social phenomena; it
provides an important tool for sifting large amounts of data to find
interesting results in the absence of hypothesis testing.
